---
title: The public opinion strategy behind Republican abortion rhetoric
author: John D. Johnson
date: '2019-05-01'
slug: the-public-opinion-strategy-behind-republican-abortion-rhetoric
categories:
  - United States
  - Public opinion
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
---


```{r setup, include=FALSE}
library(tidyverse)
library(MLSPTools)
library(haven)
library(labelled)
knitr::opts_chunk$set(echo = TRUE)
life.choice <- readxl::read_excel("~/Dropbox/Projects/2019/April/Abortion/Gallup_abortion1.xlsx") %>%
  gather(key = "position", value = "pct", - DateString, -Date) %>%
  mutate(Date = as.Date(Date))
quinnipiac <- readxl::read_excel("~/Dropbox/Projects/2019/April/Abortion/Quinnipiac_abortion.xlsx") %>%
  gather(key = "position", value = "pct", - DateString, -Date) %>%
  mutate(Date = as.Date(Date))
integ <- read_stata("~/Dropbox/MuLawPoll1/Waves/2019/52_April/Data/integ52.dta") %>%
  filter(zwave > 45)
```

Republicans appear to be making abortion a central political issue in the runup to the 2020 elections. While abortion policy is a salient issue in nearly every state and federal election, I believe that Republicans are pursuing a new tactic in this election cycle. Republicans in Congress and various state legislatures including Virginia and Wisconsin have introduced legislation requiring doctors to save the lives of babies who are inadvertently born during the course of an abortion procedure. Despite the fact that this legislation apparently describes a situation which does not exist, President Trump has lent his vocal and graphic support to these efforts.[^1] In the rest of this post, I'll discuss the political strategy behind these bills.[^2]

Most people know what they think about abortion. Polls routinely find just 5% or fewer are unwilling or unable to give an opinion.[^3] And if it's framed as pro-life vs. pro-choice binary, the population seems evenly and oppositely divided.

```{r, echo=FALSE, fig.cap="Source: Gallup Poll", fig.align='center', fig.height=3, message=FALSE}
ggplot(life.choice, aes(Date, pct, col = position, group = position)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(labels = scales::percent_format(scale = 1),
                     name = NULL) +
  scale_x_date(name = NULL) +
  labs(title = "With respect to the abortion issue,\nwould you consider yourself to be pro-choice or pro-life?") +
  ggthemes::theme_tufte() +
  theme(legend.title = element_blank())
```

In reality, views of abortion are on a continuum. At one extreme is the bill proposed by Texas legislator Tony Tinderholt which would allow for women who have abortions to be executed for first degree murder.[^4] At the other extreme are the views of people like bioethicist Peter Singer who finds infanticide acceptable as he believes newborns lack "rationality, automony, and self-consciousness."[^5] Nearly everyone lies somewhere between these two poles. (The range of options presented by American politicians clearly skews right. After all, Tinderholt represents a state legislative district and no elected officials that I'm aware of advocate infanticide.[^6])

A more nuanced question gives people a scale of 4 options--always legal, legal in most cases, illegal in most cases, or always illegal. When given these choices, some movement in public opinion over time is evident. Compared to 15 years ago, more people think abortion should be "legal in most cases," and fewer think it should be illegal in most or all cases.

```{r, echo=FALSE, fig.cap="Source:  Quinnipiac University Poll", fig.align='center', fig.height=3, message=FALSE}
ggplot(quinnipiac, aes(Date, pct, col = position, group = position)) +
  geom_point() +
  geom_smooth() +
  scale_y_continuous(labels = scales::percent_format(scale = 1),
                     name = NULL) +
  scale_x_date(name = NULL) +
  labs(title = "Do you think abortion should be legal in all cases,\nlegal in most cases, illegal in most cases or illegal in all cases?") +
  ggthemes::theme_tufte() +
  theme(legend.title = element_blank())
```

Wisconsin was the state decided by the smallest margin in 2016, and there is a very good chance it will be just as divided in 2020. Its views on abortion closely mirror the nation as a whole, so it makes a good case study. Here are the attitudes of Wisconsin registered voters in 2018, according to the Marquette Law Poll.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Source: Marquette Law Poll", fig.height=3.5}
integ %>%
  filter(!is.na(h83)) %>%
  mutate(total = sum(zwave_weight),
         response = to_factor(h83)) %>%
  group_by(response) %>%
  summarise(pct = (sum(zwave_weight)/first(total))*100) %>%
  ggplot(aes(response, pct)) +
  geom_bar(stat = "identity") +
  geom_label(aes(response, pct, label = round(pct, digits = 0)),
            family = "serif", vjust = 1) +
  scale_y_continuous(name = NULL, labels = scales::percent_format(scale = 1)) +
  scale_x_discrete(name = NULL) +
  labs(title = "Do you think abortion should be legal in all cases,\nlegal in most cases, illegal in most cases, or illegal in all cases?") +
  ggthemes::theme_tufte() +
  theme(axis.text.x = element_text(angle = -40, hjust = 0))
```

You can spin these numbers in several directions. *A majority think abortion should have minimal restrictions* **and** *Two-thirds of Wisconsinites think abortion should have legal restrictions* are both technically accurate interpretations of this data. But clearly there is not majority support for either of the "extreme" positions. The parties have to assemble their working majority by drawing from the middle categories.

Usually, the dynamics of the Republican party primary force the winning candidate toward the right. Because every candidate on the debate stage is (almost always) pro-life, the debate becomes a contest to prove one's anti-abortion bona fides. At stake is the support of the powerful strongly pro-life faction of Republican primary voters. In appealing to them, however, the Republican candidate risks alienating the large share of voters who hold more ambivalent positions.

In recent years, state-level Republican parties have generally pursued their anti-abortion aims along two tracks.

1. They have passed legislation designed to make operating an abortion clinic prohibitively difficult without directly challenging *Roe v. Wade*. These efforts include legislating the width of hallways or requiring physicians at abortion clinics to have admitting privileges at local hospitals. While such laws are often effective, they aren't particularly useful as a political rallying cry outside a narrow (if important) base of highly attuned pro-life activists.
2. In a more blunt tactic, Republicans in some state legislatures (including Kansas, Oklahoma, and Ohio) have passed legislation criminalizing abortion altogether.[^7] This is plainly in direct conflict with *Roe*, and even the supporters of these bills generally consider them quixotic statements. Still, they send a strong signal to all pro-life constituents that these politicians mean business.

Neither of these tactics helps Republicans expand their coalition of voters beyond the third of the American electorate who actually support strict anti-abortion policies. This leaves Republican strategists in a bind. If they moderate on abortion policies they risk being primaried at worst or simply losing the enthusiasm of pro-life voters at best. The Constitution Party already challenges Republicans by describing itself as the "only 100% pro-life party."

Bill Clinton was surely aware of this dynamic when he characterized his administration's position on abortion as ensuring that is was "safe, legal, and rare."[^8] I don't know how many Americans were positively enthusiastic about this formulation, but it staked out a consensus position that a solid majority of voters could at least accept. When a quarter of Americans support unrestricted abortion access and another third support only limited restrictions, a strictly pro-life Republican faces an uphill battle.

That's why, I believe, Republicans are introducing these new "accidental birth" (or whatever you want to call it) laws. They can't moderate for fear of losing their base, so the policy positions they represent are perpetually stuck in the minority. Instead, they are trying to use these new bills to paint Democrats as being on the extreme left of the abortion issue. I believe this tactic is designed for two purposes.

1. Perhaps Republicans can peel off some voters with more middling views on abortion. 38% of Independents who lean Democratic think abortion should be legal in "most cases" and 13% think it should be "illegal in most cases." In a narrowly divided state like Wisconsin any movement here can be critical.
2. Just as importantly, I think Republicans are introducing this legislation to shore up support among their own traditional base. White evangelicals are famously supportive of Trump, but their enthusiasm may be shakier than commonly thought.[^9] In Wisconsin, just 50% of white evangelicals say they would "definitely" vote for Trump if the election was today. Another 18% say they would "probably" vote for Trump; similarly, 20% of white evangelicals only "somewhat" support the president. For the subset of evangelicals who are genuinely troubled by aspects of Donald Trump's policies and personality, provocative portrayals of pro-choice politicans may be an effective way to bring them back "home" to the Republican party. If that is impossible, it may still make them sufficiently disgusted with the Democrats that they wont vote for either major party, which is the next best option for the G.O.P.


[^1]: https://www.nytimes.com/2019/04/29/opinion/trump-wisconsin-abortion.html
[^2]: I do not mean to suggest that abortion is merely a "political issue." More importantly, it is a healthcare and human rights issue. But since this legislation is not designed to pass or to remedy an actually existing situation (so far as I can tell), I think it is appropriate to describe as a political gambit.
[^3]: http://pollingreport.com/abortion.htm
[^4]: https://www.texastribune.org/2019/04/11/texas-lawmaker-abortion-bill-leach-tinderholt-safety-concern/
[^5]: https://en.wikipedia.org/wiki/Peter_Singer#Abortion,_euthanasia,_and_infanticide
[^6]: with the possible past exception of former California Senator Barabara Boxer. https://www.newsweek.com/george-will-barbara-boxers-position-abortion-74293
[^7]: https://kfor.com/2018/11/30/oklahoma-senator-proposing-bill-to-abolish-abortion-in-oklahoma-calls-on-state-to-ignore-federal-rulings/; https://radio.wosu.org/post/ohio-legislature-considers-total-abortion-ban#stream/0; https://www.tulsaworld.com/news/capitol_report/sen-silk-files-bill-to-make-abortion-illegal-classifying-it/article_83a6c39b-05e2-5a93-985a-8340acb8efff.html
[^8]: http://www.ontheissues.org/Celeb/Bill_Clinton_Abortion.htm
[^9]: By "evangelical" I mean anyone who self-describes as a "born-again or evangelical" Christian.