---
title: Most common degrees by occupation
author: John D. Johnson
date: '2020-09-16'
slug: most-common-degrees-by-occupation
categories:
  - Jobs
  - United States
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
output: 
  html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(reactable)
d <- readRDS("/Users/johnsonjoh/Dropbox/Projects/2020/September/GenderIncome/data/DegreesAndOccupations.rds") %>%
  select(occupation = OCC2010, "pct w/degree" = 2,
         "most common_degree" = 3, "most common_pct" = 6,
         "2nd most common_degree" = 4, "2nd most common_pct" = 7,
         "3rd most common_degree" = 5, "3rd most common_pct" = 8)
```

This table shows the share of workers in each occupational category with a bachelor's degree, along with the top three most common degrees for each job. I built the table using over 7 million interviews from the 2014-18 American Community Survey, but occupations with very small numbers of college degree holders nonetheless have large margins of error for the most common college degrees.

Data source:

> Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version 10.0 [dataset]. Minneapolis, MN: IPUMS, 2020.
https://doi.org/10.18128/D010.V10.0

```{r, echo=FALSE}
reactable(d, searchable = TRUE, filterable = TRUE, striped = TRUE,
          columns = list(
            "pct w/degree" = colDef(format = colFormat(suffix = "%", digits = 0)),
            "most common_pct" = colDef(format = colFormat(suffix = "%", digits = 0),
                                       name = "pct"),
            "2nd most common_pct" = colDef(format = colFormat(suffix = "%", digits = 0),
                                           name = "pct"),
            "3rd most common_pct" = colDef(format = colFormat(suffix = "%", digits = 0),
                                           name = "pct"),
            "most common_degree" = colDef(name = "degree"),
            "2nd most common_degree" = colDef(name = "degree"),
            "3rd most common_degree" = colDef(name = "degree")
          ),
          columnGroups = list(
            colGroup(name = "most common degree", 
                     columns = c("most common_degree", "most common_pct")),
            colGroup(name = "2nd most common degree", 
                     columns = c("2nd most common_degree", "2nd most common_pct")),
            colGroup(name = "3rd most common degree", 
                     columns = c("3rd most common_degree", "3rd most common_pct"))
          ))

```
