---
title: Neighborhoods where Milwaukee isn't segregated
author: John D. Johnson
date: '2020-06-17'
slug: milwaukee-neighborhoods-race-2018
categories:
  - Milwaukee
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(reactable)
library(leaflet)
library(sf)

d <- read_rds("~/Dropbox/Projects/2020/June/MKE_neighborhoods_race/MilwaukeeNeighborhoodsByRace.rds")

plurality <- d %>%
  group_by(neighborhood) %>%
  filter(pct == max(pct)) %>%
  mutate(plurality = race)

majority <- plurality %>%
  mutate(plurality = replace(plurality, pct <= 50, "no majority")) %>%
  rename(majority = plurality)

plurality %>% group_by(race) %>% summarise(neighborhoods = n())
majority %>% group_by(race) %>% summarise(neighborhoods = n())
no.majority <- majority %>% filter(majority == "no majority") %>%
  arrange(desc(total_pop)) %>%
  pull(neighborhood) %>%
  str_to_title()

d2 <- inner_join(d, majority %>% select(neighborhood, majority)) %>%
  inner_join(plurality %>% select(neighborhood, plurality, plurality_pct = pct)) %>%
  select(-pop) %>%
  mutate(race = str_to_title(race)) %>%
  pivot_wider(names_from = race, values_from = pct) %>%
  mutate(`All others` = (100 - Asian - Black - Latino - White)) %>%
  ungroup() %>%
  mutate(label = paste0('<p>', htmltools::htmlEscape(str_to_title(neighborhood)), '<p></p>', 
                         htmltools::htmlEscape(paste("Total pop:", prettyNum(round(total_pop),
                                                                             big.mark = ","))), '</p><p>',
                         htmltools::htmlEscape(paste("majority:", str_to_title(majority))), '</p><p>',
                         htmltools::htmlEscape(paste(paste0(round(Asian),"%"), "Asian")),'</p><p>',
                         htmltools::htmlEscape(paste(paste0(round(Black),"%"), "Black")),'</p><p>',
                         htmltools::htmlEscape(paste(paste0(round(Latino),"%"), "Latino")),'</p><p>',
                         htmltools::htmlEscape(paste(paste0(round(White),"%"), "White")),'</p><p>',
                         htmltools::htmlEscape(paste(paste0(round(`All others`),"%"), "Others")), '</p>'))

nhood.shape <- st_read("~/Dropbox/Projects/SHPfiles/neighborhood/neighborhood.shp") %>%
  select(neighborhood = NEIGHBORHD) %>%
  inner_join(d2) %>%
  st_transform(crs = 4326)
```

Milwaukee's pervasive segregation is infamous, but it doesn't extend everywhere. [Brown University’s American Communities Project](https://s4.ad.brown.edu/Projects/Diversity/SegSorting/Default.aspx) ranked the metro in 2010 as 2nd in Black/white segregation, 13th in Latino/white segregation, and 116th in Asian/white segregation. The metro area includes all of Milwaukee County plus neighboring, Ozaukee, Washington, and Waukesha counties, but the situation is scarcely better [within city limits](https://s4.ad.brown.edu/Projects/Diversity/SegCitySorting/Default.aspx). In fact it's worse for Latino/white and Asian/white segregation. The City of Milwaukee ranked 10th in Black/white segregation, 8th in Latino/white segregation, and 16th in Asian/white segregation.

Still, not all of Milwaukee's neighborhoods are characterized by racial segregation. Among [Milwaukee's 186 neighborhoods](https://city.milwaukee.gov/Neighborhoods), 73 have a Black majority, 72 have a white majority, 14 are majority Latino, and 27 have no majority group. I calculated these statistics by (dis)aggregating data from the 2018 5-year American Community Survey into neighborhood boundaries using [the methodology described here](https://github.com/jdjohn215/Milwaukee-Geo-Crosswalks). Similarily, Piggsville (also called "The Valley") has 36% Black residents, 34% white, 16% Latino, and 11% Asian.

The 27 neighborhoods without a majority group are, arranged in descending order by total population, `r paste(no.majority)`.

The most integrated neighborhoods are Little Menomonee Parkway (on the Northwest side) and Piggsville (near Miller Park). The residents of Little Menomonee Parkway are 35% Black, 35% white, 15% Latino, 8% something else, and 7% Asian.

The statistics for each neighborhood are presented in the table below.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
d %>%
  filter(race != "native") %>%
  mutate(race = str_to_title(race),
         neighborhood = str_to_title(neighborhood),
         total_pop = round(total_pop)) %>%
  select(-pop) %>%
  pivot_wider(names_from = race, values_from = pct) %>%
  mutate(`All others` = (100 - Asian - Black - Latino - White)) %>%
  mutate_if(is.numeric, round) %>%
  rename("Total pop." = 2,
         "Neighborhood" = 1) %>%
  reactable(striped = TRUE,
            showPageSizeOptions = TRUE,
            filterable = TRUE,
            sortable = TRUE,
            columns = list(
              `Total pop.` = colDef(format = colFormat(separators = TRUE)),
              Asian = colDef(format = colFormat(suffix = "%")),
              Black = colDef(format = colFormat(suffix = "%")),
              Latino = colDef(format = colFormat(suffix = "%")),
              White = colDef(format = colFormat(suffix = "%")),
              `All others` = colDef(format = colFormat(suffix = "%"))
            ))
```

Move your mouse over this map to see the statistics from the table above.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=11, fig.align='center'}
green.pal <- colorNumeric(
  palette = colorRampPalette(colors = c("#bae4b3", "#006d2c"))(length(d2$plurality_pct)),
  domain = d2$plurality_pct
)

blue.pal <- colorNumeric(
  palette = colorRampPalette(colors = c("#bdd7e7", "#08519c"))(length(d2$plurality_pct)),
  domain = d2$plurality_pct
)

purple.pal <- colorNumeric(
  palette = colorRampPalette(colors = c("#cbc9e2", "#54278f"))(length(d2$plurality_pct)),
  domain = d2$plurality_pct
)

red.pal <- colorNumeric(
  palette = colorRampPalette(colors = c("#fcae91", "#a50f15"))(length(d2$plurality_pct)),
  domain = d2$plurality_pct
)

leaflet() %>%
  setView(lat = 43.06, lng = -87.96, zoom = 12) %>%
  addProviderTiles("Stamen.TonerLite") %>%
  addPolygons(
    data = nhood.shape %>% filter(majority == "black"),
    fillColor = ~green.pal(plurality_pct),
    weight = 0,
    fillOpacity = 0.5,
    label = ~lapply(label, htmltools::HTML),
    popup = ~lapply(label, htmltools::HTML),
    highlightOptions = highlightOptions(fillOpacity = 0.7,
                                        bringToFront = TRUE)
  ) %>%
  addPolygons(
    data = nhood.shape %>% filter(majority == "white"),
    fillColor = ~blue.pal(plurality_pct),
    weight = 0,
    fillOpacity = 0.5,
    label = ~lapply(label, htmltools::HTML),
    popup = ~lapply(label, htmltools::HTML),
    highlightOptions = highlightOptions(fillOpacity = 0.7,
                                        bringToFront = TRUE)
  ) %>%
  addPolygons(
    data = nhood.shape %>% filter(majority == "latino"),
    fillColor = ~purple.pal(plurality_pct),
    weight = 0,
    fillOpacity = 0.5,
    label = ~lapply(label, htmltools::HTML),
    popup = ~lapply(label, htmltools::HTML),
    highlightOptions = highlightOptions(fillOpacity = 0.7,
                                        bringToFront = TRUE)
  ) %>%
  addPolygons(
    data = nhood.shape %>% filter(majority == "no majority"),
    fillColor = ~red.pal(plurality_pct),
    weight = 0,
    fillOpacity = 0.5,
    label = ~lapply(label, htmltools::HTML),
    popup = ~lapply(label, htmltools::HTML),
    highlightOptions = highlightOptions(fillOpacity = 0.7,
                                        bringToFront = TRUE)
  )

```
