---
title: Retrieving precipitation data from NOAA
author: John D. Johnson
date: '2020-06-05'
slug: retrieving-precipitation-data-from-noaa
categories:
  - United States
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
---

This post details how to retrieve the daily precipitation data needed to create the graph [demonstrated here](https://johndjohnson.info/post/how-to-build-a-tufte-style-weather-graph-in-r-using-ggplot2/).

## Getting the data
This graph requires 3 datasets, all of them from a specific NOAA weather station.

* current year (2020) daily precipitation counts
* climatological normal year-to-date precipitation
* daily record high and low year-to-date precipitation

First, choose a weather station which has been around long enough for NOAA to calculate climatological normals. [Download the full list as a .txt file](ftp://ftp.ncdc.noaa.gov/pub/data/normals/1981-2010/station-inventories/allstations.txt). The first and last columns contain unique station IDs.

I'm using the NOAA station at Milwaukee's primary airport, General Mitchell Intl (MKE). The station code is `USW00014839` and the district code is `726400`. 

### current year weather reports

All of the 2020 daily weather reports are available as individual files for each station in the directory `ftp://ftp.ncdc.noaa.gov/pub/data/gsod/2020/`. The URL for the MKE station is `ftp://ftp.ncdc.noaa.gov/pub/data/gsod/2020/726400-14839-2020.op.gz`. 

This code downloads the data:

```{r get2020, message=FALSE}
library(tidyverse)

general.mitchell.2020 <- read_table(file = "ftp://ftp.ncdc.noaa.gov/pub/data/gsod/2020/726400-14839-2020.op.gz") %>%
  # just select station_id, date, and precipitation
  select(station_id = 1, year_month_day = 3, precip = PRCP) %>%
  separate(year_month_day, sep = c(4,6), into = c("year","month","day")) %>%
  mutate(
    # this removes the data quality flag
    precip = str_remove(precip, "G|I|H"),
    # values of 99.99 indicate missing value, replace with 0
    precip = as.numeric(replace(precip, precip == 99.99, 0)),
    # create cumulative precipitation
    cum_precip_2020 = cumsum(precip),
    # convert column classes
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day),
    station_id = as.character(station_id),
    # create date variable
    date = as.Date(paste(year, month, day, sep = "-"))) %>%
  # just keep the variables we want
  select(month, day, cum_precip_2020)
head(general.mitchell.2020)
```

### climatological normals

This code downloads the year-to-date precipitation normals for each weather station. After downloading, we filter for just the Milwaukee station.

```{r getNormals, message=FALSE}
precip.normals <- read_table("ftp://ftp.ncdc.noaa.gov/pub/data/normals/1981-2010/products/precipitation/ytd-prcp-normal.txt",
                             col_names = c("station_id", "month", paste(1:31))) %>%
  # pivot so that there is one row for each date
  pivot_longer(names_to = "day", values_to = "precip",
               cols = -c(station_id, month)) %>%
  # extract the quality flag from end of precip (details in documentation)
  mutate(quality_flag = str_sub(precip, -1),
         # replace flag with NA when there is no letter flag
         quality_flag = replace(quality_flag, ! quality_flag %in% c("C","P","Q","R","S"), NA),
         # remove flags from precipitation
         precip = str_remove(precip, "C|P|Q|R|S"),
         # convert classes
         day = as.numeric(day),
         month = as.numeric(month),
         precip = as.numeric(precip)) %>%
  # remove dates that don't exists
  filter(precip != -8888) %>%
  # convert precipitation to inches
  mutate(precip = precip/100,
         precip = replace(precip, precip < 0, NA),
         # replace cumprecip is missing with the previous day's value
         precip = zoo::na.locf(precip)) %>%
  # filter for the desired station
  filter(station_id == "USW00014839") %>%
  select(-quality_flag)
head(precip.normals)
```

### records

As far as I know, NOAA doesn't provide a dataset with station records, so I had to calculate this myself. You could download each annual file from NOAA's FTP server by changing the year in the file path above, or you could use NOAA's [data download user interface](https://www.ncdc.noaa.gov/cdo-web/datatools/lcd). Once you've acquired the data, calculate the cumulative year-to-date precipitation for each day, then group by day of year. Calculate the standard deviation using just the years 1981-2010, because this is the 30-year period used to calculate the climatological normals retrieved above. Your output should look something like this. You can also download this file using the provided dropbox link.

```{r getRecords, message=FALSE}
precip.records <- read_rds(url("https://www.dropbox.com/s/62w9jtx5inae3l8/MKE_precip_records.rds?dl=1"))
head(precip.records)
```

### combine the data

Join the three `data.frames` by month and day, then create a day-of-year variable.

```{r joinData, message=FALSE}
all.precip.data <- inner_join(precip.normals, precip.records) %>%
  left_join(general.mitchell.2020) %>%
  mutate(dayofleapyear = 1:length(station_id))
head(all.precip.data)
```
