---
title: Changes in Wisconsin voter registration from October 1 to November 1, 2020
author: John D. Johnson
date: '2020-11-02'
slug: changes-in-wisconsin-voter-registration-from-october-1-to-november-1-2020
categories:
  - Wisconsin
  - Elections
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(reactable)
library(gt)

counties <- read_rds("/Users/johnsonjoh/Dropbox/Projects/2020/October/EarlyVoting/RV/RV_Oct_Nov_Counties.rds")
municipalities <- read_rds("/Users/johnsonjoh/Dropbox/Projects/2020/October/EarlyVoting/RV/RV_Oct_Nov_Municipalities.rds")
wards <- read_rds("/Users/johnsonjoh/Dropbox/Projects/2020/October/EarlyVoting/RV/RV_Oct_Nov_Wards.rds")
age<- read_rds("/Users/johnsonjoh/Dropbox/Projects/2020/October/EarlyVoting/RV/RV_Oct_Nov_Age.rds")
```

The following tables compare voter registration totals from the Wisconsin Election Commission on [October 1st](https://elections.wi.gov/node/7147) and [November 1st](https://elections.wi.gov/node/7220) 2020.

## County registered voter totals
```{r}
counties %>%
  mutate(county = str_to_title(county)) %>%
  select(county, october, november, change, pct_change, `voting age citizens` = vep, `nov pct of voting age citizens` = nov_pct_of_vep) %>%
  reactable(filterable = T,
            columns = list(
              october = colDef(format = colFormat(separators = T)),
              november = colDef(format = colFormat(separators = T)),
              change = colDef(format = colFormat(separators = T)),
              `voting age citizens` = colDef(format = colFormat(separators = T)),
              pct_change = colDef(format = colFormat(suffix = "%", digits = 1)),
              `nov pct of voting age citizens` = colDef(format = colFormat(suffix = "%", digits = 1))
            ))
```

## Municipality registered voter totals
```{r}
municipalities %>%
  mutate(county = str_to_title(county),
         municipality = str_to_title(municipality)) %>%
  select(county, municipality, october, november, change, pct_change) %>%
  reactable(filterable = T,
            columns = list(
              october = colDef(format = colFormat(separators = T)),
              november = colDef(format = colFormat(separators = T)),
              change = colDef(format = colFormat(separators = T)),
              pct_change = colDef(format = colFormat(suffix = "%", digits = 1))
            ))
```


## Ward registered voter totals
```{r}
wards %>%
  mutate(county = str_to_title(county)) %>%
  select(county, ward, october, november, change, pct_change) %>%
  reactable(filterable = T,
            columns = list(
              october = colDef(format = colFormat(separators = T)),
              november = colDef(format = colFormat(separators = T)),
              change = colDef(format = colFormat(separators = T)),
              pct_change = colDef(format = colFormat(suffix = "%", digits = 1))
            ))
```


## Age group registered voter totals
```{r}
age %>%
  select(age_group, october, november, change, pct_change) %>%
  reactable(filterable = T,
            columns = list(
              october = colDef(format = colFormat(separators = T)),
              november = colDef(format = colFormat(separators = T)),
              change = colDef(format = colFormat(separators = T)),
              pct_change = colDef(format = colFormat(suffix = "%", digits = 1))
            ))
```


*Eagle-eyed readers may notice that the total number of registered voters is not exactly identical between the files. I believe this is because the WEC's files are created from a "live" system continuously receiving updates from local clerks. Consequently, data for individual records may be updated between the time it takes to run, e.g., the county and ward summary scripts.*
